---
title: "ordination & diversity metrics"
author: "Judy-Malas"
date: "12/22/2021"
output: html_document
---

## Before Starting 

* Make sure your clear the global environment by clicking on the broom in R studio 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
  
### Load packages
```{r load-packages, message=FALSE, warning=FALSE}

library(ggplot2)
library(gridExtra)
library(phyloseq)
require("tidyverse")
library("dplyr")
library("rmarkdown")
library("vegan")
library("PoiClaClu")
library("doParallel")
library("plotly")
library("microbiome")
library("ggpubr")
library("viridis")
library("plyr")
library("magrittr")
library("scales")
library(microbiome) # data analysis and visualisation
library(phyloseq) # also the basis of data object. Data analysis and visualisation
library(RColorBrewer) # nice color options
library(ggpubr) # publication quality figures, based on ggplot2
library(DT) # interactive tables in html and markdown
library(dplyr) # data handling  


```


```{r import-all-objects}

# live samples 

ps_all.step2 <- readRDS("~/landfill-microcosms/data/live_samples_objects/ps_all.step2") 

ps_all.rmout <- readRDS("~/landfill-microcosms/data/live_samples_objects/ps_all.rmout") #outlier removed

vst_all.step2 <- readRDS("~/landfill-microcosms/data/live_samples_objects/vst_all.step2")

ps_all.rl <- readRDS("~/landfill-microcosms/data/live_samples_objects/ps_all.rl") 

ps_all.prop <- readRDS("~/landfill-microcosms/data/live_samples_objects/ps_all.prop") 

ps_all.5 <- readRDS("~/landfill-microcosms/data/live_samples_objects/ps_all.5") #low prevelance taxa removed, untransformed



 #killed samples 

ps_killed.step2 <- readRDS("~/landfill-microcosms/data/killed_microcosms_objects/ps_killed.step2")

ps_killed.rmout <- readRDS("~/landfill-microcosms/data/killed_microcosms_objects/ps_killed.rmout")

vst_killed.step2 <- readRDS("~/landfill-microcosms/data/killed_microcosms_objects/vst_killed.step2")

ps_killed.rl <- readRDS("~/landfill-microcosms/data/killed_microcosms_objects/ps_killed.rl")

ps_killed.prop <- readRDS("~/landfill-microcosms/data/killed_microcosms_objects/ps_killed.prop")


```


A] Distance Measures; Ordination plots
B] Alpha diveristy
C] Taxa summaries
  C-1] Boxplots 
    0.4] make a dataframe from a phyloseq object.
  C-1-1] Arrange boxplots in a grid together
  C-2] Barcharts

-----

#### update the metadata tables

```{r add_new_column, eval=FALSE}

ps_all.step2_samdata <- ps_all.step2@sam_data #extract the sample data 

new_sam_data <- ps_all.step2_samdata %>% 
  mutate(treatment =  #add a new column of data which specifies whether the samples have ACTUALLY been treated with a spike
           case_when(sample_time == "T1" ~ "None", 
                     sample_time == "T2" ~ "None", 
                    sample_time == "T3" ~ "None", 
                        sample_time == "T4" ~ "None", 
                        sample_time == "T5" ~ as.character(spike), 
                    sample_time == "T6" ~ as.character(spike))) 

new_sam_data <- new_sam_data %>% 
  mutate(treatment = ifelse(treatment == "Control", "None", treatment)) #change "Control" to "none" in the new treatment variable


as.factor(new_sam_data$treatment)

new_sam_data <- new_sam_data %>% 
  mutate(treatment = fct_relevel(treatment, "None", "Antibiotics", "Fe(OH)3", "Na2SO4"))
  


ps_all.step2@sam_data = new_sam_data # assign the new dataframe to the phyloseq object
vst_all.step2@sam_data = new_sam_data




```


####  Assigning plot colors


These are the original colors I was using, but they are not colorblind friendly
```{r assign-colors}

# exper_colors <-  c( "gray50", "blue", "firebrick1", "gold")



```

```{r list-all-cb-friendly-palettes}

brewer.pal.info %>% 
  filter(colorblind == T)
  
```

```{r visualize-all-cb-friendly-palettes}
display.brewer.all(colorblindFriendly = T)
```


Ended up using colors in the 8-color plalette for color blindness: 

http://mkweb.bcgsc.ca/colorblind/img/colorblindness.palettes.v11.pdf

```{r updated-assign-colors}
exper_colors <- c("black", "#3DB7E9",  "#D55E00", "#F0E442")
```




#### re-simplifying names
```{r}

ps_all <- ps_all.step2
vst_all <- vst_all.step2

```


#A] Distance Measures; Ordination plots

```{r nmds-bray-vst-all}

nmds_bray_vst_all <- ordinate(vst_all, method="NMDS", distance = "bray")

nmds_bray_vst_all_plot  <-
  plot_ordination(vst_all, nmds_bray_vst_all, color = "spike", shape ="sample_time", title = "Bray-Curtis NMDS")+ 
    scale_color_manual(values= exper_colors) + 
    geom_point(size=3.5)
nmds_bray_vst_all_plot #print the plot


```

Here we're using the new treatment variable we made to make the plot
```{r nmds-bray-vst-all-treatment}

nmds_bray_vst_all_plot  <-
  plot_ordination(vst_all, nmds_bray_vst_all, color = "treatment", shape ="sample_time", title = "Bray-Curtis NMDS")+ 
    scale_color_manual(values= exper_colors) + 
    geom_point(size=3.5)+ 
    theme_bw()


nmds_bray_vst_all_plot #print the plot


```


```{r pcoa-wunifrac-vst-all}

PCoA_wunifrac_vst_all <-  phyloseq::ordinate(vst_all, method="PCoA", distance = "wunifrac")

PCoA_wunifrac_vst_all_plot  <-
  plot_ordination(vst_all , PCoA_wunifrac_vst_all, color = "spike", shape ="sample_time", title = "wUnifrac PCoA") +
  geom_point(size=3.5) +
  scale_color_manual(values= exper_colors)

PCoA_wunifrac_vst_all_plot


```

```{r pcoa-wunifrac-vst-all-treatment}

PCoA_wunifrac_vst_all_plot  <-
  plot_ordination(vst_all , PCoA_wunifrac_vst_all, color = "treatment", shape ="sample_time", title = "wUnifrac PCoA") +
  geom_point(size=3.5) +
  scale_color_manual(values= exper_colors) + 
  theme_bw()

PCoA_wunifrac_vst_all_plot

```


```{r pcoa-unifrac-untransformed}


PCoA_unifrac_vst_all <-  ordinate(ps_all, method="PCoA", distance = "unifrac")

PCoA_unifrac_vst_all_plot <-  plot_ordination(ps_all, PCoA_unifrac_vst_all, color = "treatment", shape ="sample_time", title = "unifrac PCoA Untransformed") +
    scale_color_manual(values= exper_colors) +
    geom_point(size=3.5)

PCoA_unifrac_vst_all_plot


```


#A-1] Create an ellipse around
This part doesn't work because there are too few points
```{r}

PCoA_wunifrac_vst_all_plot +
  scale_color_manual(values= exper_colors) +
  geom_point(size=3.5) +
  stat_ellipse(type="norm", geom="polygon", alpha=1/10)





```


#B] Alpha diversity 
Used to identify within individual taxa richness and evenness
The commonly used metrics/indices are Shannon, Inverse Simpson, Simpson, Gini, Observed and Chao1. These indices do not take into account the phylogeny of the taxa identified in sequencing. Phylogenetic diversity (Faith’s PD) uses phylogenetic distance to calculate the diversity of a given sample. (Sudarshan et al., 2020)


One has to consider the sequencing depth (how much of the taxa have been sampled) for each sample. If there is a large difference, then it is important to normalize the samples to equal sampling depth. First, we look at the sampling depth (no. of reads per sample).
```{r}



print(vst_all)

summary(sample_sums(vst_all)) #number of reads per sample
#there isn't a large difference between the min and max 
#I'm going to look at the non vst phyloseq object 


all <- ps_all.step2

summary(sample_sums(all))
#so obviously the vst trims down the number of reads significantly somehow. Again, I need to read the paper

#we can plot the rarefaction curve for the observed ASVs in the entire dataset, this is a way to check how the richness captured in the sequencing effort 

#otu_tab = t(abundances(vst_all)) 


otu_tab = t(abundances(all))  #t = transpose I think, not sure why we transposed it

#this step doesn't work with the vst_all, probably because of something abotu the way we transformed it


#plot of species/reads
#we're nto doing anything with this, but it's good to know
p = vegan::rarecurve(otu_tab, 
                      step = 50, label = FALSE, 
                      sample = min(rowSums(otu_tab), 
                                   col = "blue", cex = 0.6))

#the tutorial I'm using here used this plot to rarify their data, however, we're not doing this here because susan holmes would be mad.

```

Non-phylogenetic diversities (non vst)
```{r}
#make sure you have the most up-to-date version of the microbiome package; use #BiocManager::install("microbiome") #update all
#sometimes you have to unlaod and re-load the package for this step to work 
#let's calculate diversity
#this also doesn't work with vst_all
#turns out it's not suppose to work with the variance stabilizing transformation, so just use the original phyloseq object
div = alpha(all, index="all")  
data.table(div)

# get the metadata out as seprate object
meta <- meta(all)

# Add the rownames as a new colum for easy integration later.
meta$sam_name <- rownames(meta)

# Add the rownames to diversity table
div$sam_name <- rownames(div)

# merge these two data frames into one
div.df <- merge(div, meta, by = "sam_name")

# check the tables
colnames(div)


# Now use this data frame to plot 
p_div <- ggboxplot(div.df, 
               x = "spike", 
               y = "diversity_shannon",
              fill = "spike", 
              palette = "jco")

p_div <- p_div + rotate_x_text()

print(p_div)



#Seperate by both spike and time using metadata
p_div_1 <- ggboxplot(div.df, 
               x = "spike_time", 
               y = "diversity_shannon",
              fill = "sample_time")


p_div_1 <- p_div_1 + rotate_x_text()

print(p_div_1)

#I need to re-order the samples so that ABT4 is 




```


shannon phylogenetic diversity w/re-ordered samples (non vst)
```{r}
#Seperate by both spike and time using metadata

#order_of_samples = data.frame(name=c("ABT1", "ABT2", "ABT3", "ABT4" , "ABT5" ,"ABT6" ,"ABT1", "ABT2" ,
 #                                    "ABT3","ABT5", "ABT6" ,"ABT1", "ABT2", "ABT3",
  #                                   "ABT5", "ABT6" ,"FeT1","FeT2" ,"FeT3" ,"FeT4",
   #                                  "FeT5" ,"FeT6" ,"FeT4", "FeT1" ,"FeT2" ,"FeT3" ,
    #                               "FeT5" ,"FeT6", "FeT1" ,"FeT2" ,"FeT3", "FeT4" ,
     #                                "FeT5", "FeT6" ,"NLT1", "NLT2", "NLT3", "NLT4",
      #                               "NLT5", "NLT6" ,"NLT1", "NLT2" ,"NLT3" ,"NLT4",
       #                              "NLT5" ,"NLT6", "NLT1", "NLT2" ,"NLT3", "NLT4" ,
        #                             "NLT5" ,"NLT6" ,"ST1" , "ST2" , "ST3" , "ST4",
         #                            "ST5" , "ST6" , "ST1" , "ST2" , "ST3" , "ST4" , 
          #                           "ST5",  "ST6",  "ST1" , "ST2",  "ST3",  "ST4", 
           #                          "ST5" , "ST6"))

#ignore, but I'm keeping the order just in case!! 

as.factor(div.df$spike_time)

div.df$spike_time = factor(div.df$spike_time, levels = c("ABT1", "ABT2", "ABT3", "ABT4" , "ABT5" ,"ABT6", "FeT1","FeT2" ,"FeT3" ,"FeT4", "FeT5" ,"FeT6", "NLT1", "NLT2", "NLT3", "NLT4", "NLT5", "NLT6", "ST1" , "ST2",  "ST3",  "ST4", "ST5" , "ST6"))


p_div_1 <- ggboxplot(div.df, 
               x = "spike_time", 
               y = "diversity_shannon",
              fill = "sample_time")


p_div_1 <- p_div_1 + rotate_x_text()

print(p_div_1)

#THAT WORKED, IT'S ORDERED NOW!! ONLY TOOK 1 HOUR TO FIGURE OUT. 


```



simpson boxplots (non vst)
```{r}
#spike time
p_even <- ggboxplot(div.df, 
               x = "spike_time", 
               y = "evenness_simpson",
              fill = "sample_time")


p_even <- p_even + rotate_x_text()

print(p_even)

```



Shannon and Simpson (vst)
```{r}


Shannon_plot = plot_richness(vst_all.ordered, x="spike_time", measures=c("Shannon"), color="spike")
Shannon_plot + scale_color_manual(values= exper_colors) 





Simpson_plot = plot_richness(vst_all, x="spike_time", measures=c("Simpson"), color="spike")
Simpson_plot  + scale_color_manual(values= exper_colors)
```


Phylogenetic diversity using the Picante packate 
https://mibwurrepo.github.io/Microbial-bioinformatics-introductory-course-Material-2018/alpha-diversities.html#phylogenetic-diversity
all of my diveristy measures were the same, I give up on this for now
```{r, include=FALSE}

library(picante)

vst_all_asvtab = as.data.frame(vst_all@otu_table) #otu tab as a data frame
view(vst_all_asvtab)

vst_all_tree = vst_all@phy_tree #seperating the phylogenetic tree 


# We first need to check if the tree is rooted or not 

vst_all@phy_tree #unrooted tree 

#df.pd <- pd(t(vst_all_asvtab), vst_all_tree, include.root=T) # t(ou_table) transposes the table for use in picante and the tre file comes from the first code chunck we used to read tree file 

#need to root the tree, using joey711/phyloseq comment from Feb 28, 2018, issue #597 
#https://github.com/joey711/phyloseq/issues/597

pick_new_outgroup <- function(tree.unrooted){
    require("magrittr")
    require("data.table")
    require("ape") # ape::Ntip
    # tablify parts of tree that we need.
    treeDT <- 
      cbind(
        data.table(tree.unrooted$edge),
        data.table(length = tree.unrooted$edge.length)
      )[1:Ntip(tree.unrooted)] %>% 
      cbind(data.table(id = tree.unrooted$tip.label))
    # Take the longest terminal branch as outgroup
    new.outgroup <- treeDT[which.max(length)]$id
    return(new.outgroup)
}


#It returns the string of the tip-label for the new proposed outgroup. It is then up to the user/process to take that result and do the next rooting step. This is a bit easier to test and maintain than also doing the rooting and opaquely returning the tree.

new.outgroup = pick_new_outgroup(vst_all_tree)
rootedTree= ape::root(vst_all_tree, outgroup=new.outgroup, resolve.root=TRUE)

rootedTree_vst_all = rootedTree
rootedTree_vst_all


#try again to the the phylogentic diversity 

#df.pd <- pd(t(vst_all_asvtab), rootedTree,include.root=T)
## Error in UseMethod("is.rooted") : no applicable method for 'is.rooted' applied to an object of class "NULL"

#okay I'll try to make my phyloseq object again with the rooted tree and then repeat 


my_vst_physeqobj_rooted = phyloseq(seq_count_phy2, Metadata_tab_phy2, tax_tab_phy2, phy_tree(rootedTree))

my_vst_physeqobj_full_time_rooted = subset_samples(my_vst_physeqobj_rooted, bottle!="NL1" &  bottle!="NL2"& bottle!="NL3")

vst_all_rooted = my_vst_physeqobj_full_time_rooted #just renaming it to be a simplier name
 vst_all@sam_data[["bottle"]] #just to check 
 
vst_all_rooted_asvtab = as.data.frame(vst_all_rooted@otu_table)
vst_all_rooted_tree = vst_all_rooted@phy_tree


#df.pd = pd(vst_all_rooted_asvtab, vst_all_rooted_tree) #all i needed to do was NOT transpose the tree like they suggested. ugh.
#datatable(df.pd) #why is pd all the same? 

#let's try transposing first and then doing the df.pd
vst_all_rooted_asvtab.t = t(vst_all_rooted_asvtab)

df.pd = pd(vst_all_rooted_asvtab.t, vst_all_rooted_tree)

vst_all_rooted_asvtab.t


# get the metadata out as seprate object
vst_meta <- meta(vst_all_rooted)


vst_meta$Phylogenetic_Diversity = df.pd$PD #added the phylogenetic diversity results to the vst_metadata

view(vst_meta)





```






#0.3]taxa glom
```{r}
#summes all ASVs that were in the same level

#Glom at famiily level 
vst_all_rel_glom = tax_glom(vst_all_rel_1, taxrank = "Family")
vst_all_rel_glom_family = vst_all_rel_glom

#Glom at phylum level 
vst_all_rel_glom_phylum = tax_glom(vst_all_rel_1, taxrank= "Phylum")



```


#C-1] Boxplots 
```{r}

#Phylum level box plot 

vst_all_rel_glom2 = tax_glom(vst_all_rel_1, taxrank = "Phylum")

dat_vst_all_rel2 = data.table(psmelt(vst_all_rel_glom2))
dat_vst_all_rel2$Phylum <- as.character(dat_vst_all_rel2$Phylum)
dat_vst_all_rel2$Order = as.character(dat_vst_all_rel2$Order)
dat_vst_all_rel2$Family = as.character(dat_vst_all_rel2$Family)

# group dataframe by Family, calculate median rel. abundance
dat_vst_all_rel2[, median := median(Abundance, na.rm = TRUE), 
    by = "Phylum"]
# Change name to remainder of Family less than .1%
dat_vst_all_rel2[(median <= 0.001), Phylum := " Other < .1%"]


boxplot_Phylum = ggplot(dat_vst_all_rel2[Abundance > 0],
       aes(x=Phylum,
           y=Abundance)) + 
  geom_boxplot() + 
  coord_flip() +
  scale_y_log10()+ labs(title="Phylum")
boxplot_Phylum


#Family level box plot

vst_all_rel_glom = tax_glom(vst_all_rel_1, taxrank = "Family")

dat_vst_all_rel = data.table(psmelt(vst_all_rel_glom))
dat_vst_all_rel$Phylum <- as.character(dat_vst_all_rel$Phylum)
dat_vst_all_rel$Order = as.character(dat_vst_all_rel$Order)
dat_vst_all_rel$Family = as.character(dat_vst_all_rel$Family)

# group dataframe by Family, calculate median rel. abundance
dat_vst_all_rel[, median := median(Abundance, na.rm = TRUE), 
    by = "Family"]
# Change name to remainder of Family less than 5%
dat_vst_all_rel[(median <= 0.01), Family := " Other < 1%"]


boxplot_Family = ggplot(dat_vst_all_rel[Abundance > 0],
       aes(x=Family,
           y=Abundance)) + 
  geom_boxplot() + 
  coord_flip() +
  scale_y_log10()+ labs(title="Family")
boxplot_Family



#genus level boxplot
vst_all_rel_glom_genus = tax_glom(vst_all_rel_1, taxrank = "Genus")

dat_vst_all_rel_genus = data.table(psmelt(vst_all_rel_glom_genus)) #make a data table with the agg taxa at genus level 
dat_vst_all_rel_genus$Phylum <- as.character(dat_vst_all_rel_genus$Phylum)
dat_vst_all_rel_genus$Order = as.character(dat_vst_all_rel_genus$Order)
dat_vst_all_rel_genus$Family = as.character(dat_vst_all_rel_genus$Family)
dat_vst_all_rel_genus$Genus = as.character(dat_vst_all_rel_genus$Genus)



# group dataframe by Genus, calculate median rel. abundance
dat_vst_all_rel_genus[, median := median(Abundance, na.rm = TRUE), 
    by = "Genus"]
# Change name to remainder of Family less than 1%
dat_vst_all_rel_genus[(median <= 0.01), Genus := " Other < 1%"]


boxplot_genus = ggplot(dat_vst_all_rel_genus[Abundance > 0],
       aes(x=Genus,
           y=Abundance)) + 
  geom_boxplot() + 
  coord_flip() +
  scale_y_log10()+ labs(title="Genus")
boxplot_genus


```

#C-1-1] Arrange boxplots in a grid together
```{r}
ggarrange(boxplot_Phylum, boxplot_Family, nrow=1)
boxplot_genus

```

#0.4] Make a dataframe from a phyloseq object
```{r}

dat_family = data.table(psmelt(vst_all_rel_glom_family))
# convert Phylum to a character vector from a factor because R
dat_family$Phylum <- as.character(dat$Phylum)
dat_family$Order = as.character(dat$Order)
dat_family$Family = as.character(dat$Family)

# group dataframe by Family, calculate median rel. abundance
dat_family[, median := median(Abundance, na.rm = TRUE), 
    by = "Family"]
# Change name to remainder of Family less than 1%
dat_family[(median <= 0.01), Family := " Other < 1%"]
```


#C-2] Barcharts 
phylum
```{r}

# Create a data table for ggplot
vst_all_phylum <- vst_all %>%
  tax_glom(taxrank = "Phylum") %>%                     # agglomerate at phylum level
  transform_sample_counts(function(x) {x/sum(x)} ) %>% # Transform to rel. abundance (or use ps0.ra)
  psmelt() %>%                                         # Melt to long format for easy ggploting
  filter(Abundance > 0.01)                             # Filter out low abundance taxa

# Plot - Phylum
p.ra.phylum <- ggplot(vst_all_phylum, aes(x = spike_time, y = Abundance, fill = Phylum)) + 
  geom_bar(stat = "identity", width = 1) +
  facet_wrap(spike~spike_time, scales = "free_x", nrow = 4, ncol = 7) +
  theme(axis.text.x = element_blank()) +
  theme(axis.title.x = element_blank()) +
  labs(title = "Abundant Phylum (> 1%)")
p.ra.phylum


# Create a data table for ggplot
vst_all_phylum <- vst_all %>%
  tax_glom(taxrank = "Phylum") %>%                     # agglomerate at phylum level
  transform_sample_counts(function(x) {x/sum(x)} ) %>% # Transform to rel. abundance (or use ps0.ra)
  psmelt() %>%                                         # Melt to long format for easy ggploting
  filter(Abundance > 0.01)                             # Filter out low abundance taxa

# Plot - Phylum
p.ra.phylum <- ggplot(vst_all_phylum, aes(x = sample_time, y = Abundance, fill = Phylum)) + 
  geom_bar(stat = "identity", width = 1) +
  labs(title = "Abundant Phylum (> 1%)")
p.ra.phylum

mean(vst_all_phylum$Abundance~sample_time)

abundance_by_time = vst_all_phylum %>% group_by(sample_time)
levels(abundance_by_time$sample_time)
as.factor(abundance_by_time$sample_time)

rel_abundance_T1 = vst_all_phylum %>% subset(sample_time == "T1")




```
Genus
```{r}
# Create a data table for ggplot
vst_all_genus <- vst_all %>%
  tax_glom(taxrank = "Genus") %>%                     # agglomerate at phylum level
  psmelt()                                        # Melt to long format for easy ggploting
     
  
vst_all_genus_rel <- vst_all %>%
  tax_glom(taxrank="Genus") %>%
  transform_sample_counts(function(x) {x/sum(x)}) %>% # Transform to rel
  psmelt()
                                        
  
vst_all_genus$relative_abundance = vst_all_genus_rel$Abundance

vst_all_genus_filt = vst_all_genus %>%filter(relative_abundance>0.01)
vst_all_genus_filt_5 = vst_all_genus %>%filter(relative_abundance>0.05)



# Plot - Genus
p.ra.genus <- ggplot(vst_all_genus_filt_5, aes(x = sample_time, y = Abundance, fill = Genus)) + 
  geom_bar(aes(color=Genus, fill=Genus), stat = "identity", width = 1) +
  labs(title = "Abundant Genus ( >5 %)", x = "Sample Time")
p.ra.genus


# Plot - Genus
p.ra.genus_2 <- ggplot(vst_all_genus_filt_5, aes(x = spike, y = Abundance, fill = Genus)) + 
  geom_bar(aes(color=Genus, fill=Genus), stat = "identity", width = 1) +
  labs(title = "Abundant Genus (>5 %)", x="Spike")
p.ra.genus_2

#plot genus 

p.ra.genus_3 <- ggplot(vst_all_genus_filt, aes(x = bottle , y = relative_abundance, fill = Genus)) +
  geom_bar(aes(color=Genus, fill=Genus), stat = "identity", width = 1) +
  labs(title = "Abundant Genus (>1 %)", x="Spike")+
  facet_wrap(sample_time~spike, scales="free_x", nrow=4)
p.ra.genus_3


```

