---
title: "ordination & diversity metrics"
author: "Judy-Malas"
date: "12/22/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
  
### Load packages
```{r load-packages, message=FALSE, warning=FALSE}

library(ggplot2)
library(gridExtra)
library(phyloseq)
require("tidyverse")
library("dplyr")
library("rmarkdown")
library("vegan")
library("PoiClaClu")
library("doParallel")
library("plotly")
library("microbiome")
library("ggpubr")
library("viridis")
library("plyr")
library("magrittr")
library("scales")


```

```{r}
ps_all <- readRDS("~/landfill-microcosms/data/live_samples_objects/ps_all") # live samples 
vst_all <- readRDS("~/landfill-microcosms/data/live_samples_objects/vst_all") # live samples with transformed count table 
ps_all.5 <- readRDS("~/landfill-microcosms/data/live_samples_objects/ps_all.5") #low prevelance taxa removed, untransformed

ps_killed <-readRDS("~/landfill-microcosms/data/killed_microcosms_objects/ps_killed") #killed samples 
vst_killed <-readRDS("~/landfill-microcosms/data/killed_microcosms_objects/vst_killed") #killed samples with transformed count table 


```


#                          PART III: Let's make figures!   
    --> using the variance stabilizing transformation (vst) phyloseq object from here on out [except for the diversity plots e.g. shannon, simpson, etc.]
              
    0.1]Assigning plot colors 
    0.2] Transform data into relative counts
A] Distance Measures; Ordination plots
B] Alpha diveristy
C] Taxa summaries
  C-1] Boxplots 
    0.4] make a dataframe from a phyloseq object.
  C-1-1] Arrange boxplots in a grid together
  C-2] Barcharts

-----

#0.1] Assigning plot colors
```{r}

exper_colors.ordered = c( "gray50", "blue", "firebrick1", "gold")

```

#0.2] Transform data into relative counts
```{r}
vst_all_rel_count = sweep(vst_trans_count_tab.abs, 1, rowSums(vst_trans_count_tab.abs), '/')

#OR 

vst_all_rel_1 = transform_sample_counts(vst_all, function(x) {x/sum(x)}) #from joey711/phyloseq issue #494

```

#A] Distance Measures; Ordination plots
```{r}

#NMDS + Bray curtis distances 

nmds_bray_vst_all = ordinate(vst_all, method="NMDS", distance = "bray")

nmds_bray_vst_all_plot = 
  plot_ordination(vst_all, nmds_bray_vst_all, color = "spike", shape ="sample_time", title = "Bray-Curtis NMDS")+ 
    scale_color_manual(values= exper_colors) + 
    geom_point(size=3.5)
nmds_bray_vst_all_plot #print the plot


#PCoA + Unifrac 

PCoA_wunifrac_vst_all = ordinate(vst_all, method="PCoA", distance = "wunifrac")

PCoA_wunifrac_vst_all_plot = 
  plot_ordination(vst_all , PCoA_wunifrac_vst_all, color = "spike", shape ="sample_time", title = "wUnifrac PCoA") +
  geom_point(size=3.5) +
  scale_color_manual(values= exper_colors)
PCoA_wunifrac_vst_all_plot

  
#PCoA + Unifrac with the ordered phyloseq object
PCoA_wunifrac_vst_all.ordered = ordinate(vst_all, method="PCoA", distance = "wunifrac")

PCoA_wunifrac_vst_all_plot.ordered = 
  plot_ordination(vst_all.ordered, PCoA_wunifrac_vst_all.ordered, color = "spike", shape ="sample_time", title = "wunifrac PCoA Ordered") +
    scale_color_manual(values= exper_colors) +
    geom_point(size=3.5)
PCoA_wunifrac_vst_all_plot.ordered



# so for some reason ordered in the phyloseq object with control first doesn't put the legend in order that the phyloseq object is, the only thing that happened is that the plot rotated 360 degrees.. I kinda like the ordered one better, it's easier to see somehow

```
#A-1] Create an ellipse around
This part doesn't work because there are too few points
```{r}

PCoA_wunifrac_vst_all_plot.ordered + scale_color_manual(values= exper_colors) + geom_point(size=3.5) +
stat_ellipse(type="norm", geom="polygon", alpha=1/10, aes (fills=sample_time))
# too few points
print(PCoA_wunifrac_vst_all.ordered +stat_ellipse())


```


#B] Alpha diversity 
Used to identify within individual taxa richness and evenness
The commonly used metrics/indices are Shannon, Inverse Simpson, Simpson, Gini, Observed and Chao1. These indices do not take into account the phylogeny of the taxa identified in sequencing. Phylogenetic diversity (Faith’s PD) uses phylogenetic distance to calculate the diversity of a given sample. (Sudarshan et al., 2020)


One has to consider the sequencing depth (how much of the taxa have been sampled) for each sample. If there is a large difference, then it is important to normalize the samples to equal sampling depth. First, we look at the sampling depth (no. of reads per sample).
```{r}
library(microbiome) # data analysis and visualisation
library(phyloseq) # also the basis of data object. Data analysis and visualisation
library(RColorBrewer) # nice color options
library(ggpubr) # publication quality figures, based on ggplot2
library(DT) # interactive tables in html and markdown
library(dplyr) # data handling  


print(vst_all)

summary(sample_sums(vst_all)) #number of reads per sample
#there isn't a large difference between the min and max 
#I'm going to look at the non vst phyloseq object 

summary(sample_sums(all))
#so obviously the vst trims down the number of reads significantly somehow. Again, I need to read the paper

#we can plot the rarefaction curve for the observed ASVs in the entire dataset, this is a way to check how the richness captured in the sequencing effort 

#otu_tab = t(abundances(vst_all)) 


otu_tab = t(abundances(all))  #t = transpose I think, not sure why we transposed it

#this step doesn't work with the vst_all, probably because of something abotu the way we transformed it


#plot of species/reads
#we're nto doing anything with this, but it's good to know
p = vegan::rarecurve(otu_tab, 
                      step = 50, label = FALSE, 
                      sample = min(rowSums(otu_tab), 
                                   col = "blue", cex = 0.6))

#the tutorial I'm using here used this plot to rarify their data, however, we're not doing this here because susan holmes would be mad.

```

Non-phylogenetic diversities (non vst)
```{r}
#make sure you have the most up-to-date version of the microbiome package; use #BiocManager::install("microbiome") #update all
#sometimes you have to unlaod and re-load the package for this step to work 
#let's calculate diversity
#this also doesn't work with vst_all
#turns out it's not suppose to work with the variance stabilizing transformation, so just use the original phyloseq object
div = alpha(all, index="all")  
data.table(div)

# get the metadata out as seprate object
meta <- meta(all)

# Add the rownames as a new colum for easy integration later.
meta$sam_name <- rownames(meta)

# Add the rownames to diversity table
div$sam_name <- rownames(div)

# merge these two data frames into one
div.df <- merge(div, meta, by = "sam_name")

# check the tables
colnames(div)


# Now use this data frame to plot 
p_div <- ggboxplot(div.df, 
               x = "spike", 
               y = "diversity_shannon",
              fill = "spike", 
              palette = "jco")

p_div <- p_div + rotate_x_text()

print(p_div)



#Seperate by both spike and time using metadata
p_div_1 <- ggboxplot(div.df, 
               x = "spike_time", 
               y = "diversity_shannon",
              fill = "sample_time")


p_div_1 <- p_div_1 + rotate_x_text()

print(p_div_1)

#I need to re-order the samples so that ABT4 is 




```


shannon phylogenetic diversity w/re-ordered samples (non vst)
```{r}
#Seperate by both spike and time using metadata

#order_of_samples = data.frame(name=c("ABT1", "ABT2", "ABT3", "ABT4" , "ABT5" ,"ABT6" ,"ABT1", "ABT2" ,
 #                                    "ABT3","ABT5", "ABT6" ,"ABT1", "ABT2", "ABT3",
  #                                   "ABT5", "ABT6" ,"FeT1","FeT2" ,"FeT3" ,"FeT4",
   #                                  "FeT5" ,"FeT6" ,"FeT4", "FeT1" ,"FeT2" ,"FeT3" ,
    #                               "FeT5" ,"FeT6", "FeT1" ,"FeT2" ,"FeT3", "FeT4" ,
     #                                "FeT5", "FeT6" ,"NLT1", "NLT2", "NLT3", "NLT4",
      #                               "NLT5", "NLT6" ,"NLT1", "NLT2" ,"NLT3" ,"NLT4",
       #                              "NLT5" ,"NLT6", "NLT1", "NLT2" ,"NLT3", "NLT4" ,
        #                             "NLT5" ,"NLT6" ,"ST1" , "ST2" , "ST3" , "ST4",
         #                            "ST5" , "ST6" , "ST1" , "ST2" , "ST3" , "ST4" , 
          #                           "ST5",  "ST6",  "ST1" , "ST2",  "ST3",  "ST4", 
           #                          "ST5" , "ST6"))

#ignore, but I'm keeping the order just in case!! 

as.factor(div.df$spike_time)

div.df$spike_time = factor(div.df$spike_time, levels = c("ABT1", "ABT2", "ABT3", "ABT4" , "ABT5" ,"ABT6", "FeT1","FeT2" ,"FeT3" ,"FeT4", "FeT5" ,"FeT6", "NLT1", "NLT2", "NLT3", "NLT4", "NLT5", "NLT6", "ST1" , "ST2",  "ST3",  "ST4", "ST5" , "ST6"))


p_div_1 <- ggboxplot(div.df, 
               x = "spike_time", 
               y = "diversity_shannon",
              fill = "sample_time")


p_div_1 <- p_div_1 + rotate_x_text()

print(p_div_1)

#THAT WORKED, IT'S ORDERED NOW!! ONLY TOOK 1 HOUR TO FIGURE OUT. 


```



simpson boxplots (non vst)
```{r}
#spike time
p_even <- ggboxplot(div.df, 
               x = "spike_time", 
               y = "evenness_simpson",
              fill = "sample_time")


p_even <- p_even + rotate_x_text()

print(p_even)

```



Shannon and Simpson (vst)
```{r}


Shannon_plot = plot_richness(vst_all.ordered, x="spike_time", measures=c("Shannon"), color="spike")
Shannon_plot + scale_color_manual(values= exper_colors) 





Simpson_plot = plot_richness(vst_all, x="spike_time", measures=c("Simpson"), color="spike")
Simpson_plot  + scale_color_manual(values= exper_colors)
```


Phylogenetic diversity using the Picante packate 
https://mibwurrepo.github.io/Microbial-bioinformatics-introductory-course-Material-2018/alpha-diversities.html#phylogenetic-diversity
all of my diveristy measures were the same, I give up on this for now
```{r, include=FALSE}

library(picante)

vst_all_asvtab = as.data.frame(vst_all@otu_table) #otu tab as a data frame
view(vst_all_asvtab)

vst_all_tree = vst_all@phy_tree #seperating the phylogenetic tree 


# We first need to check if the tree is rooted or not 

vst_all@phy_tree #unrooted tree 

#df.pd <- pd(t(vst_all_asvtab), vst_all_tree, include.root=T) # t(ou_table) transposes the table for use in picante and the tre file comes from the first code chunck we used to read tree file 

#need to root the tree, using joey711/phyloseq comment from Feb 28, 2018, issue #597 
#https://github.com/joey711/phyloseq/issues/597

pick_new_outgroup <- function(tree.unrooted){
    require("magrittr")
    require("data.table")
    require("ape") # ape::Ntip
    # tablify parts of tree that we need.
    treeDT <- 
      cbind(
        data.table(tree.unrooted$edge),
        data.table(length = tree.unrooted$edge.length)
      )[1:Ntip(tree.unrooted)] %>% 
      cbind(data.table(id = tree.unrooted$tip.label))
    # Take the longest terminal branch as outgroup
    new.outgroup <- treeDT[which.max(length)]$id
    return(new.outgroup)
}


#It returns the string of the tip-label for the new proposed outgroup. It is then up to the user/process to take that result and do the next rooting step. This is a bit easier to test and maintain than also doing the rooting and opaquely returning the tree.

new.outgroup = pick_new_outgroup(vst_all_tree)
rootedTree= ape::root(vst_all_tree, outgroup=new.outgroup, resolve.root=TRUE)

rootedTree_vst_all = rootedTree
rootedTree_vst_all


#try again to the the phylogentic diversity 

#df.pd <- pd(t(vst_all_asvtab), rootedTree,include.root=T)
## Error in UseMethod("is.rooted") : no applicable method for 'is.rooted' applied to an object of class "NULL"

#okay I'll try to make my phyloseq object again with the rooted tree and then repeat 


my_vst_physeqobj_rooted = phyloseq(seq_count_phy2, Metadata_tab_phy2, tax_tab_phy2, phy_tree(rootedTree))

my_vst_physeqobj_full_time_rooted = subset_samples(my_vst_physeqobj_rooted, bottle!="NL1" &  bottle!="NL2"& bottle!="NL3")

vst_all_rooted = my_vst_physeqobj_full_time_rooted #just renaming it to be a simplier name
 vst_all@sam_data[["bottle"]] #just to check 
 
vst_all_rooted_asvtab = as.data.frame(vst_all_rooted@otu_table)
vst_all_rooted_tree = vst_all_rooted@phy_tree


#df.pd = pd(vst_all_rooted_asvtab, vst_all_rooted_tree) #all i needed to do was NOT transpose the tree like they suggested. ugh.
#datatable(df.pd) #why is pd all the same? 

#let's try transposing first and then doing the df.pd
vst_all_rooted_asvtab.t = t(vst_all_rooted_asvtab)

df.pd = pd(vst_all_rooted_asvtab.t, vst_all_rooted_tree)

vst_all_rooted_asvtab.t


# get the metadata out as seprate object
vst_meta <- meta(vst_all_rooted)


vst_meta$Phylogenetic_Diversity = df.pd$PD #added the phylogenetic diversity results to the vst_metadata

view(vst_meta)





```






#0.3]taxa glom
```{r}
#summes all ASVs that were in the same level

#Glom at famiily level 
vst_all_rel_glom = tax_glom(vst_all_rel_1, taxrank = "Family")
vst_all_rel_glom_family = vst_all_rel_glom

#Glom at phylum level 
vst_all_rel_glom_phylum = tax_glom(vst_all_rel_1, taxrank= "Phylum")



```


#C-1] Boxplots 
```{r}

#Phylum level box plot 

vst_all_rel_glom2 = tax_glom(vst_all_rel_1, taxrank = "Phylum")

dat_vst_all_rel2 = data.table(psmelt(vst_all_rel_glom2))
dat_vst_all_rel2$Phylum <- as.character(dat_vst_all_rel2$Phylum)
dat_vst_all_rel2$Order = as.character(dat_vst_all_rel2$Order)
dat_vst_all_rel2$Family = as.character(dat_vst_all_rel2$Family)

# group dataframe by Family, calculate median rel. abundance
dat_vst_all_rel2[, median := median(Abundance, na.rm = TRUE), 
    by = "Phylum"]
# Change name to remainder of Family less than .1%
dat_vst_all_rel2[(median <= 0.001), Phylum := " Other < .1%"]


boxplot_Phylum = ggplot(dat_vst_all_rel2[Abundance > 0],
       aes(x=Phylum,
           y=Abundance)) + 
  geom_boxplot() + 
  coord_flip() +
  scale_y_log10()+ labs(title="Phylum")
boxplot_Phylum


#Family level box plot

vst_all_rel_glom = tax_glom(vst_all_rel_1, taxrank = "Family")

dat_vst_all_rel = data.table(psmelt(vst_all_rel_glom))
dat_vst_all_rel$Phylum <- as.character(dat_vst_all_rel$Phylum)
dat_vst_all_rel$Order = as.character(dat_vst_all_rel$Order)
dat_vst_all_rel$Family = as.character(dat_vst_all_rel$Family)

# group dataframe by Family, calculate median rel. abundance
dat_vst_all_rel[, median := median(Abundance, na.rm = TRUE), 
    by = "Family"]
# Change name to remainder of Family less than 5%
dat_vst_all_rel[(median <= 0.01), Family := " Other < 1%"]


boxplot_Family = ggplot(dat_vst_all_rel[Abundance > 0],
       aes(x=Family,
           y=Abundance)) + 
  geom_boxplot() + 
  coord_flip() +
  scale_y_log10()+ labs(title="Family")
boxplot_Family



#genus level boxplot
vst_all_rel_glom_genus = tax_glom(vst_all_rel_1, taxrank = "Genus")

dat_vst_all_rel_genus = data.table(psmelt(vst_all_rel_glom_genus)) #make a data table with the agg taxa at genus level 
dat_vst_all_rel_genus$Phylum <- as.character(dat_vst_all_rel_genus$Phylum)
dat_vst_all_rel_genus$Order = as.character(dat_vst_all_rel_genus$Order)
dat_vst_all_rel_genus$Family = as.character(dat_vst_all_rel_genus$Family)
dat_vst_all_rel_genus$Genus = as.character(dat_vst_all_rel_genus$Genus)



# group dataframe by Genus, calculate median rel. abundance
dat_vst_all_rel_genus[, median := median(Abundance, na.rm = TRUE), 
    by = "Genus"]
# Change name to remainder of Family less than 1%
dat_vst_all_rel_genus[(median <= 0.01), Genus := " Other < 1%"]


boxplot_genus = ggplot(dat_vst_all_rel_genus[Abundance > 0],
       aes(x=Genus,
           y=Abundance)) + 
  geom_boxplot() + 
  coord_flip() +
  scale_y_log10()+ labs(title="Genus")
boxplot_genus


```

#C-1-1] Arrange boxplots in a grid together
```{r}
ggarrange(boxplot_Phylum, boxplot_Family, nrow=1)
boxplot_genus

```

#0.4] Make a dataframe from a phyloseq object
```{r}

dat_family = data.table(psmelt(vst_all_rel_glom_family))
# convert Phylum to a character vector from a factor because R
dat_family$Phylum <- as.character(dat$Phylum)
dat_family$Order = as.character(dat$Order)
dat_family$Family = as.character(dat$Family)

# group dataframe by Family, calculate median rel. abundance
dat_family[, median := median(Abundance, na.rm = TRUE), 
    by = "Family"]
# Change name to remainder of Family less than 1%
dat_family[(median <= 0.01), Family := " Other < 1%"]
```


#C-2] Barcharts 
phylum
```{r}

# Create a data table for ggplot
vst_all_phylum <- vst_all %>%
  tax_glom(taxrank = "Phylum") %>%                     # agglomerate at phylum level
  transform_sample_counts(function(x) {x/sum(x)} ) %>% # Transform to rel. abundance (or use ps0.ra)
  psmelt() %>%                                         # Melt to long format for easy ggploting
  filter(Abundance > 0.01)                             # Filter out low abundance taxa

# Plot - Phylum
p.ra.phylum <- ggplot(vst_all_phylum, aes(x = spike_time, y = Abundance, fill = Phylum)) + 
  geom_bar(stat = "identity", width = 1) +
  facet_wrap(spike~spike_time, scales = "free_x", nrow = 4, ncol = 7) +
  theme(axis.text.x = element_blank()) +
  theme(axis.title.x = element_blank()) +
  labs(title = "Abundant Phylum (> 1%)")
p.ra.phylum


# Create a data table for ggplot
vst_all_phylum <- vst_all %>%
  tax_glom(taxrank = "Phylum") %>%                     # agglomerate at phylum level
  transform_sample_counts(function(x) {x/sum(x)} ) %>% # Transform to rel. abundance (or use ps0.ra)
  psmelt() %>%                                         # Melt to long format for easy ggploting
  filter(Abundance > 0.01)                             # Filter out low abundance taxa

# Plot - Phylum
p.ra.phylum <- ggplot(vst_all_phylum, aes(x = sample_time, y = Abundance, fill = Phylum)) + 
  geom_bar(stat = "identity", width = 1) +
  labs(title = "Abundant Phylum (> 1%)")
p.ra.phylum

mean(vst_all_phylum$Abundance~sample_time)

abundance_by_time = vst_all_phylum %>% group_by(sample_time)
levels(abundance_by_time$sample_time)
as.factor(abundance_by_time$sample_time)

rel_abundance_T1 = vst_all_phylum %>% subset(sample_time == "T1")




```
Genus
```{r}
# Create a data table for ggplot
vst_all_genus <- vst_all %>%
  tax_glom(taxrank = "Genus") %>%                     # agglomerate at phylum level
  psmelt()                                        # Melt to long format for easy ggploting
     
  
vst_all_genus_rel <- vst_all %>%
  tax_glom(taxrank="Genus") %>%
  transform_sample_counts(function(x) {x/sum(x)}) %>% # Transform to rel
  psmelt()
                                        
  
vst_all_genus$relative_abundance = vst_all_genus_rel$Abundance

vst_all_genus_filt = vst_all_genus %>%filter(relative_abundance>0.01)
vst_all_genus_filt_5 = vst_all_genus %>%filter(relative_abundance>0.05)



# Plot - Genus
p.ra.genus <- ggplot(vst_all_genus_filt_5, aes(x = sample_time, y = Abundance, fill = Genus)) + 
  geom_bar(aes(color=Genus, fill=Genus), stat = "identity", width = 1) +
  labs(title = "Abundant Genus ( >5 %)", x = "Sample Time")
p.ra.genus


# Plot - Genus
p.ra.genus_2 <- ggplot(vst_all_genus_filt_5, aes(x = spike, y = Abundance, fill = Genus)) + 
  geom_bar(aes(color=Genus, fill=Genus), stat = "identity", width = 1) +
  labs(title = "Abundant Genus (>5 %)", x="Spike")
p.ra.genus_2

#plot genus 

p.ra.genus_3 <- ggplot(vst_all_genus_filt, aes(x = bottle , y = relative_abundance, fill = Genus)) +
  geom_bar(aes(color=Genus, fill=Genus), stat = "identity", width = 1) +
  labs(title = "Abundant Genus (>1 %)", x="Spike")+
  facet_wrap(sample_time~spike, scales="free_x", nrow=4)
p.ra.genus_3


```



#              Part IIII: Significance Testing and Other statistics 

0.I] subsetting data in phyloseq object  
I. ADONIS (Handley workflow)
II. Differential Abundance testing (DESEQ2) # Hasn't quite worked yet Need to read the DESEQ2 paper and workflow maybe?
III. Random Forest Modelling

Links to other methods to detect differentially abundant taxa:

Random Forest in R example: https://rpubs.com/michberr/randomforestmicrobe
LEfSe: https://bitbucket.org/biobakery/biobakery/wiki/lefse
ANCOM: https://www.ncbi.nlm.nih.gov/pubmed/26028277


----
#0.I] Subsetting data
```{r}
#Make a phyloseq object with just the post-spike time points
sample_data(vst_all)$sample_time #check what the levels look like
postspike_vst_all = subset_samples(vst_all, sample_time!="T1" & sample_time!="T2" & sample_time!="T3" & sample_time!="T4") #exclude anything before the spikes were added
sample_data(postspike_vst_all)$sample_time #check again 


#Make a phyloseq object with just the pre-spike time points
sample_data(vst_all)$sample_time #check what the levels look like
prespike_vst_all = subset_samples(vst_all, sample_time!="T5" & sample_time!="T6")
sample_data(prespike_vst_all)$sample_time #check again


#Make objects with just control and treament group
sample_data(vst_all)$spike

control_antibiotics = subset_samples(vst_all, spike!="Fe(OH)3" & spike!="Na2SO4")
sample_data(control_antibiotics)$spike

control_iron = subset_samples(vst_all, spike!="Antibiotics" & spike!="Na2SO4")
sample_data(control_iron)$spike

control_sulfate = subset_samples(vst_all, spike!="Fe(OH)3" & spike!="Antibiotics")
sample_data(control_sulfate)$spike

```


#I. Group Signifcance Testing with ADONIS (from handely workflow) (done)
another way:
https://mibwurrepo.github.io/Microbial-bioinformatics-introductory-course-Material-2018/beta-diversity-metrics.html#phylogenetic-beta-diversity-metrics


paper source:
https://onlinelibrary.wiley.com/doi/10.1002/9781118445112.stat07841

```{r}
#Set a random seed so that exact results can be reproduced
#I honestly don't know what that means. 
set.seed(10000)


#function to run ADONIS test on a phyloseq object and a variable from metadata 
doadonis <- function(physeq, category) {
  bdist <- phyloseq::distance(physeq, "wunifrac")
  col <- as(sample_data(physeq), "data.frame")[ ,category]
  # Adonis test
  adonis.bdist <- adonis(bdist ~ col, permutations = 10000) 
  #the great number of permuations gives greater precision, but also takes longer to run, although this didn't take that long
  print("Adonis results:")
  print(adonis.bdist)}


all_timepoints_adonis = doadonis(vst_all, "spike") #Results are significant, though I'm not sure which groups are different from each other. 

all_timepoints_adonis$aov.tab$'Pr(>F)'[1] #this gives you the p value of the adonis test 
#same thing but in another way; double bracket notation; 
all_timepoints_adonis[["aov.tab"]][["Pr(>F)"]][1]


#control v. antibiotics
control_antibiotics_adonis = doadonis(control_antibiotics, "spike") #sig

#control v. iron 
control_iron_adonis = doadonis(control_iron, "spike") #sig

#control v. sulfate
control_sulfate_adonis = doadonis(control_sulfate, "spike") #sig


##adonis of samples only pre-spike 
pre_spike_adonis = doadonis(prespike_vst_all, "spike") #significant

##adonis of samples only post-spike 
post_spike_adonis = doadonis(postspike_vst_all, "spike") #significant 


#it seems like the experiments were different even before the spikes


```

#I-1. Checking the homogeneity condition? (not yet)
https://mibwurrepo.github.io/Microbial-bioinformatics-introductory-course-Material-2018/beta-diversity-metrics.html#permanova

```{r}




```




#II. Differential Abundance Testing [[This didn't work yet]]
DESEQ2
Since the example data set collected time-series data we can take advantage of the likelihood ration test (LRT) model of DESeq2. The LRT will identify ASV differentially abundant between groups at one or more time point.
Determine all taxa differentially abundant between treatment groups at at least one time point
```{r}

# Test for taxa which at one or more time points after time 0 showed a treatment-specific effect
# Convert phyloseq object to DESeq2 table
ds_control_antibiotics_LRT <- phyloseq_to_deseq2(control_antibiotics, ~spike + sample_time + spike:sample_time)

# Run DESeq2
dds_control_antibiotics_LRT <- DESeq(ds_control_antibiotics_LRT, test="LRT", reduced = ~spike + sample_time)



# Tabulate results
res.dds.vehicle_metro.LRT <- results(dds.vehicle_metro.LRT)
res.dds.vehicle_metro.LRT$symbol <- mcols(dds.vehicle_metro.LRT)$symbol
summary(res.dds.vehicle_metro.LRT)
mcols(res.dds.vehicle_metro.LRT)
write.table(res.dds.vehicle_metro.LRT, file = "../results/deseq_vehicle_v_metro_LRT.txt", sep = "\t")

nrow(res.dds.vehicle_metro.LRT)
df.res <- as.data.frame(res.dds.vehicle_metro.LRT[ which(res.dds.vehicle_metro.LRT$padj < 0.05), ])
nrow(df.res)
df.res <- rownames_to_column(df.res, var = "ASV")
write.table(df.res, file = "../results/df_deseq2_results.txt", sep = "\t")

# Create appropriately formatted taxa table
# RDP
tax.table <- as.tibble(as.data.frame(tax_table(ps2)))
tax.table <- rownames_to_column(tax.table, var = "ASV")
df.rdp <- left_join(df.res, tax.table, by = "ASV")
colnames(df.rdp)

ggplot(df.rdp, aes(x = Phylum, y = log2FoldChange, color = Family)) +
  geom_jitter(size = 3, alpha = 0.7, width = 0.1) +
  geom_hline(yintercept = 0, lty = 2) +
  ylim(-35, 35)




dds = DEseq(dds)





```





#III. Random Forest Modelling
https://rpubs.com/michberr/randomforestmicrobe
```{r}

#need to do some basic reformatting and filtering 
colnames(tax_table(vst_all))

# Filter out, chloroplasts and mitochondria k
vst_all %>%
  subset_taxa( Family != "mitochondria" & 
              Class != "Chloroplast") -> vst_all 

vst_all





```



#Core Microbiotaanalysis
https://mibwurrepo.github.io/Microbial-bioinformatics-introductory-course-Material-2018/core-microbiota.html

```{r}

vst_all_rel_1 = transform_sample_counts(vst_all, function(x) {x/sum(x)}) #from joey711/phyloseq issue #494

print(vst_all_rel_1)

vst_all_rel_pruned = prune_taxa(taxa_sums(vst_all_rel_1)>0, vst_all_rel_1)

print(vst_all_pruned) #nothing changed so not sure why we did this step 

core.taxa.standard = core_members(vst_all_rel_pruned, detection = 0.001)

print(core.taxa.standard)

taxonomy_rel = as.data.frame(tax_table(vst_all_rel_1))

#subset taxomony to include only core OTUs

core_taxa_id = subset(taxonomy_rel, rownames(taxonomy_rel) %in% core.taxa.standard) #these are the core taxa that we've identified


#next we'll do the sum of the abundances of the core members in each sample

core.abundance <- sample_sums(core(vst_all_rel_1, detection = 0.001, prevalence = 50/100))


new = prune_taxa(core_taxa_id, vst_all_rel_1)



```



#Advanced Models for differential abudnance
https://mibwurrepo.github.io/Microbial-bioinformatics-introductory-course-Material-2018/advanced-models-for-differential-abundance.html

```{r}

```



#Trying to create a count tab with meaningful rownames
incomplete, ignore
```{r}

taxonomy = as.data.frame(tax_table(vst_all)) #extracting the taxatable from vst_all phyloseq object 

counts_all = as.data.frame(otu_table(vst_all)) #extracting the otu table 
counts_all.t = t(counts_all) #transposing the otutable

taxonomy$taxalabel = paste(new_taxanames$Family, new_taxanames$Genus) #adding a column with the Family and Genus name in one label 
replace(taxonomy$taxalabel, taxonomy$taxalabel=="NA NA", "Other") #placing the "NA NA" data with "other"

taxonomy$taxalabel2 = paste(new_taxanames$Order, new_taxanames$Family) 

replace(taxonomy$taxalabel2, taxonomy$taxalabel2=="NA NA", "Other")

identical(rownames(counts_all.t), rownames(taxonomy))#are the rownames identifical?
#TRUE

rownames(taxonomy) = taxonomy$taxalabel2

summary(vst_all)




```

